---
title: "Cheese Analysis"
output:
  html_document:
    df_print: paged
---

## Goals of this Notebook

- Analyze my cheese data
- Look for trends
- Look at yearly averages
- Play with visualizations

I want to explore my data by answering the following questions:

- Which year had the highest average cheese price? Which had the lowest?
- Which year had the highest average cheese price in 2015 dollars? Which year had the lowest price in 2015 dollars?
- For each region, which year had the highest average cheese price? Which year had the lowest?
- For each region, which year had the highest average cheese price in 2015 dollars? Which year had the lowest price in 2015 dollars?
- For 2000 through 2022, which month of the year had the highest cheese prices? Which had the lowest?

## About the data

This is [USDA](http://rmarkdown.rstudio.com) price data on domestic cheeses that I've collected, cleaned and adjusted for inflation.

## Setup

```{r setup}
library(tidyverse)
library(lubridate)
library(dplyr)
```

## Import data

I'm importing my cleaned data here.

```{r import}
chs_clean <- readRDS("data-processed/01-cleaning.rds")

chs_clean |> summary() #check my columns
```

## High and low average prices

```{r avg_unadj}
chs_clean |> 
  group_by(yr) |> 
  summarize(avg_high = mean(high_price)
  ) |>
  arrange(avg_high |> desc())
```


## Find average prices

```{r}
avg_adj_prices <- chs_clean |> 
  mutate(adj_avg_price = (adj_high_price + adj_low_price)/2,
         avg_price = (low_price + high_price)/2)

avg_adj_prices
```

## Find yearly average highs and lows

I want to get the average high price and average low price for all domestic cheese by year.

```{r avg}
chs_averages <- chs_clean |> 
  group_by(yr) |> 
  summarize(average_yearly_low = mean(low_price),
            average_yearly_high = mean(high_price),
            adj_yearly_high = mean(adj_high_price),
            adj_yearly_low = mean(adj_low_price)
  ) |> 
  arrange()

chs_averages
```

## Cheddar Cheese Average Plot

```{r american}
ched_avg <- chs_clean |> 
  filter(type == "CHEDDAR")) |> 
  group_by(yr) |> 
  summarize(average_yearly_low = mean(low_price),
            average_yearly_high = mean(high_price),
            adj_yearly_high = mean(adj_high_price),
            adj_yearly_low = mean(adj_low_price)
  ) |> 
  arrange()

ched_avg
```

Now I'll plot the data.

```{r}
ggplot(ched_avg, aes(x=yr)) + 
  geom_line(aes(y=adj_yearly_high, color="High Price")) + 
  geom_line(aes(y=adj_yearly_low, color="Low Price")) +
  labs(title = "Cheddar Cheese Prices Decrease Over Time", 
       x = "Year", y = "Price of Cheddar Cheese (in 2015 dollars")
```

## Facet Wrap All Cheeses

```{r}
facet <- avg_adj_prices |> 
  group_by(yr, type) |> 
  summarize(avg_price) |> 
  group_by(yr, type) |> 
  summarize(avg_yearly_price = mean(avg_price)) |>
  arrange()

facet
```


```{r}
facet_plot <- ggplot(facet, aes(x = yr, y = avg_yearly_price)) + 
  geom_point(aes(color = type)) +
  geom_line(aes(color = type)) +
  labs(title = "Yearly Domestic Cheese Prices (in 2015 dollars)", 
       x = "Year", y = "Price of Domestic Cheese")

facet_plot
```


```{r}
separate_plot <- facet_plot +
  facet_wrap(~ type) +
  theme(legend.position = "none")

separate_plot
```

```{r}
ggsave("plot1.pdf", plot=separate_plot, width=12, height=8)
```


