---
title: "Cheese Analysis"
output:
  html_document:
    df_print: paged
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

## Goals of this Notebook

- Look for trends in data
- Answer questions and find averages
- Play with visualizations

I want to explore my data by answering the following questions:

- Which year had the highest average cheese price? Which had the lowest?
- Which year had the highest average prices in 2015 dollars? Which had the lowest?
- Which month out of the year the highest average cheese prices in 2015 dollars? I'll group by year and month, then month again to get this answer.
- What were the monthly avarage prices for each type of cheese in 2021?
- Last year, which month had the highest total cheese prices?
- Which year had the highest cheese storage holdings, on average? Which had the lowest?
- Which week in the data had the highest cheese storage holdings? Which week had the lowest?


## About the data

This is [USDA](http://rmarkdown.rstudio.com) price data on domestic cheeses that I've collected, cleaned and adjusted for inflation.

Storage holdings data is in units of pounds.

## Setup

```{r setup}
library(tidyverse)
library(lubridate)
library(dplyr)
```

## Import data

I'm importing my cleaned data here.

```{r import}
chs_clean <- readRDS("data-processed/01-cleaning.rds")
storage_clean <- readRDS("data-processed/01-storage.rds")
```

## Create average price column

Right now each row of my data has a corresponding high and low price for each week. I'll want a column that takes the average of those two prices for each week so I can get yearly averages. 

```{r avg_column}
chs_with_avg <- chs_clean |> 
  mutate(
    total_avg_price = (high_price + low_price)/2,
    total_price_adj = (adj_high_price + adj_low_price)/2
)

chs_with_avg |> glimpse()
```

## Average total prices per year

### Which year had the highest average total cheese prices, not adjusted for inflation? Which had the lowest?

I'll group by year, which will put all the rows for each year together. Then I'll use the sum(mean) function to get the average of the total price column I just created for each year.

```{r unadj_yearly}
chs_with_avg |> 
  group_by(yr) |> 
  summarize(
    total_yr_price = mean(total_avg_price)
) |>
  arrange(total_yr_price |> desc())
```

2014 had the highest average cheese price, not adjusted for inflation, with all cheese packages averaging a price of $2.81 per pound in that years dollars.

2000 had the lowest average total cheese price in our data set, which was not adjusted for inflation, with an average minimum price of $1.49 per pound.

### Which year had the highest cheese prices in 2015 dollars? Which had the lowest?

I'll use the same logic here to find the highest total adjusted prices by year, but I'll sum the adjusted column instead.

```{r adj_yearly}
chs_with_avg |> 
  group_by(yr) |> 
  summarize(
    adj_yr_price = mean(total_price_adj)
) |>
  arrange(adj_yr_price |> desc())
```

2014 still had the highest total cheese prices after adjusting for inflation. The average price was 3.21 dollars per pound in 2015 dollars. 

The second highest year changed after adjusting for inflation, from 2022 with 2.79 dollars per pound to 2008 with 3.10 dollars per pound.

2018 had the lowest total cheese prices after adjusting for inflation, with an average price of 2.35 dollars per pound.

### In which month was the price of cheese the highest, on average?

First, I need to create a month column.

```{r month_column}
chs_mnths <- chs_with_avg |> 
  mutate(month_nm = month(date, label = TRUE)) #makes a month column
```

Now, I'll group by year and month, then month again to get the monthly average for all years.

```{r monthly}
chs_mnths |> 
  group_by(yr, month_nm) |> #group by both year and month first
  summarize(
   average_yearly_price = mean(total_price_adj)
) |> 
  group_by(month_nm) |>
  summarize(
    average_monthly_price = mean(average_yearly_price) #taking the average of the yearly average for each month
) |> 
  arrange(average_monthly_price |> desc()
)
```

October saw the highest average cheese prices in our data, with an average price of 2.83 dollars per pound in 2021 dollars.

### 2021 prices

I want to look at 2021 prices (not adjusted for inflation) for each type of cheese.

```{r lastyear}
chs_mnths |>
  filter(yr == 2021) |> 
  group_by(type) |> #group by both month and type
  summarize(
   average_price = mean(total_avg_price) 
) |> 
  arrange(average_price |> desc()
)
```

Swiss was the most expensive type of cheese on average in 2021 with an average price of 3.52 dollars per pound. Process American cheese was the least expensive that year, with an average price of 1.84 dollars per pound.

Now I want to look at total 2021 prices by month. I'll filter to 2021 and then group by month.

```{r monthly_lastyear}
chs_mnths |>
  filter(yr == 2021) |> 
  group_by(month_nm) |> #group by both month and type
  summarize(
   average_monthly_price = mean(total_avg_price) 
) |> 
  arrange(average_monthly_price |> desc()
)
```

December 2021 saw the highest average total cheese price for that year, with an average price of 2.54 per pound, adjusted for inflation.

### Swiss prices by year

I want to look at just Swiss prices, so I'll filter by data to Swiss and look at means again.

```{r swiss_by_year}
chs_clean |> 
  filter(type == "SWISS") |> 
  group_by(yr) |> 
  summarize(
    yrly_adj_high_price = mean(adj_high_price),
    yrly_adj_low_price = mean(adj_low_price)
) |> 
  arrange(yrly_adj_high_price |> desc())
```

2014 had the highest average maximum swiss cheese price after adjusting for inflation, with 3.84 dollars per pound in 2015 dollars.

The average high price for swiss was around 90 cents lower in 2018, with an adjusted price of 2.95 dollars per pound.

### Muenster prices by year

```{r cheddar}
chs_clean |> 
  filter(type == "MUENSTER") |> 
  group_by(yr) |> 
  summarize(
    yrly_adj_high_price = mean(adj_high_price),
    yrly_adj_low_price = mean(adj_low_price)
) |> 
  arrange(yrly_adj_high_price |> desc())
```

2014 saw the highest average Muesnter prices after inflation.

## Storage holdings analysis

### Which year had the highest cheese storage holdings, on average? Which had the lowest?

```{r storage}
storage_clean |> 
  group_by(yr) |> 
  summarize(avg_yrly_holding = mean(holdings_current_lbs)) |> 
  arrange(avg_yrly_holding |> desc())
```

2005 had the highest average cheese storage holdings, with over 136 million pounds of cheese stored by the U.S. government. 

2020 had the lowest average storage holdings, with over 78 million.

### Which week in the data had the highest cheese storage holdings? Which week had the lowest?

```{r storage_highs}
storage_clean |> 
  arrange(holdings_current_lbs |> desc()) |> 
  head(10)
```
July 2004 saw the greatest storage holdings, with over 150 million pounds of cheese. 

## Playing with plots

### Swiss cheese plot

```{r swiss}
swiss_avg <- chs_clean |> 
  filter(str_detect(type, "SWISS")) |> 
  group_by(yr) |> 
  summarize(average_yearly_low = mean(low_price),
            average_yearly_high = mean(high_price),
            adj_yearly_high = mean(adj_high_price),
            adj_yearly_low = mean(adj_low_price)
  ) |> 
  arrange()

swiss_avg
```

Now I'll plot the data.

```{r swiss_plot}
ggplot(swiss_avg, aes(x=yr)) + 
  geom_line(aes(y=adj_yearly_high, color="High Price")) + 
  geom_line(aes(y=adj_yearly_low, color="Low Price")) +
  labs(title = "Swiss Cheese Prices Over Time", 
       x = "Year", y = "Price per pound (in 2015 dollars)")
```

### Facet Wrap All Cheeses, Adjusted

```{r adj_facet_wrp}
facet_adj <- chs_with_avg |> 
  group_by(yr, type) |> 
  summarize(total_price_adj) |> 
  group_by(yr, type) |> 
  summarize(adj_yearly_price = mean(total_price_adj)) |>
  arrange()

facet_adj
```


```{r types}
separate_plot <- ggplot(facet_adj, aes(x = yr, y = adj_yearly_price)) + 
  geom_point(aes(color = type)) +
  geom_line(aes(color = type)) +
  labs(title = "U.S. Cheese Prices by Type (in 2021 dollars)", 
       x = "Year", y = "Price of Domestic Cheese")

separate_plot
```

```{r facet_plot}
facet_plot <- separate_plot +
  facet_wrap(~ type) +
  theme(legend.position = "none")

facet_plot
```

### Facet Wrap All Cheeses, Not Adjusted

```{r facet_unadj}
unadj <- chs_with_avg |> 
  group_by(yr, type) |> 
  summarize(total_avg_price) |> 
  group_by(yr, type) |> 
  summarize(unadj_price = mean(total_avg_price)) |>
  arrange()

unadj
```

```{r}
unadj_plot <- ggplot(unadj, aes(x = yr, y = unadj_price)) + 
  geom_point(aes(color = type)) +
  geom_line(aes(color = type)) +
  labs(title = "U.S. Cheese Prices by Type (not adjusted for inflation)", 
       x = "Year", y = "Price of Domestic Cheese")

unadj_plot
```

```{r}
unadj_facet <- unadj_plot +
  facet_wrap(~ type) +
  theme(legend.position = "none")

unadj_facet
```


```{r facet_save}
ggsave("plot1.pdf", plot=separate_plot, width=12, height=8)
```

## Lede draft

Domestic U.S. cheese prices peaked in 2014, reaching an average of $2.81 per pound, according to an analysis of U.S.D.A. data that ranged from 2000 to 2022 and adjusted for inflation. The same year year, the U.S. government stored over 90 million pounds of cheese in Madison, Wisconsin.  
