---
title: "Cheese Analysis"
output:
  html_document:
    df_print: paged
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
---

## Goals of this Notebook

- Look for trends in data
- Answer questions and find averages
- Create visualizations

I want to explore my data by answering the following questions:

### Price Questions:

- Which year had the highest average cheese price? Which had the lowest?
- Which year had the highest average prices after adjusting for inflation (in 2021 dollars)? Which had the lowest?
- In which month was the price of cheese the highest, on average (inflation adjusted)?
- What were the 2021 prices for each type of cheese?
- What were the average 2021 cheese prices for each month?
- What were the average 2021 cheese prices for each month?

### Storage Questions:

- Which year had the highest cheese storage holdings, on average? Which had the lowest?
- 

### Visualizations:


## About the data

This is [USDA](http://rmarkdown.rstudio.com) price data on domestic cheeses that I've collected, cleaned and adjusted for inflation.

Storage holdings data is in units of pounds. The storage holding data comes from the USDA National Agricultural Statistics Service [Cold Storage Report](https://quickstats.nass.usda.gov/results/172E17CF-AD11-37D8-949A-0624FA6A9CD0).

## Setup

```{r setup}
library(tidyverse)
library(lubridate)
library(dplyr)
library(scales)
library(ggthemes)
```

## Import data

I'm importing my cleaned data here.

```{r import}
chs_clean <- readRDS("data-processed/01-cleaning.rds")
storage_clean <- readRDS("data-processed/01-storage.rds")
```

### Create average price column

Right now each row of my data has a corresponding high and low price for each week. I'll want a column that takes the average of those two prices for each week so I can get yearly averages. 

```{r avg_column}
chs_with_avg <- chs_clean |> 
  mutate(
    total_avg_price = (high_price + low_price)/2,
    total_price_adj = (adj_high_price + adj_low_price)/2
)

chs_with_avg |> glimpse()
```

### Create month column

I also need to create a month column.

```{r month_column}
chs_mnths <- chs_with_avg |> 
  mutate(month_nm = month(date, label = TRUE)) #makes a month column
```

## Price Data Analysis

### Which year had the highest average total cheese prices, not adjusted for inflation? Which had the lowest?

I'll group by year, which will put all the rows for each year together. Then I'll use the sum(mean) function to get the average of the total price column I just created for each year.

```{r unadj_yearly}
chs_with_avg |> 
  group_by(yr) |> 
  summarize(
    total_yr_price = mean(total_avg_price)
) |>
  arrange(total_yr_price |> desc())
```

2014 had the highest average cheese price, not adjusted for inflation, with all cheese packages averaging a price of $2.81 per pound in that years dollars.

2000 had the lowest average total cheese price in my data, which was not adjusted for inflation, with an average minimum price of $1.65 per pound.

### Which year had the highest cheese prices in 2021 dollars? Which had the lowest?

I'll use the same logic here to find the highest total adjusted prices by year, but I'll sum the adjusted column instead.

```{r adj_yearly}
chs_with_avg |> 
  group_by(yr) |> 
  summarize(
    adj_yr_price = mean(total_price_adj)
) |>
  arrange(adj_yr_price |> desc())
```

2014 still had the highest total cheese prices after adjusting for inflation. The average price was 3.21 dollars per pound in 2021 dollars. 

The second highest year changed after adjusting for inflation, from 2022 with 2.79 dollars per pound to 2008 with 3.10 dollars per pound.

2018 had the lowest total cheese prices after adjusting for inflation, with an average price of 2.35 dollars per pound.

### In which month was the price of cheese the highest, on average?

I'll group by year and month, then month again to get the monthly average for all years.

```{r monthly}
chs_mnths |> 
  group_by(yr, month_nm) |> #group by both year and month first
  summarize(
   average_yearly_price = mean(total_price_adj)
) |> 
  group_by(month_nm) |>
  summarize(
    average_monthly_price = mean(average_yearly_price) #taking the average of the yearly average for each month
) |> 
  arrange(average_monthly_price |> desc()
)
```

October saw the highest average cheese prices in our data, with an average price of 2.83 dollars per pound in 2021 dollars.

### 2021 prices

I want to look at 2021 prices for each type of cheese.

```{r lastyear}
chs_mnths |>
  filter(yr == 2021) |> 
  group_by(type) |> #group by both month and type
  summarize(
   average_price = mean(total_avg_price) 
) |> 
  arrange(average_price |> desc()
)
```

Swiss was the most expensive type of cheese on average in 2021 with an average price of 3.52 dollars per pound. Process American cheese was the least expensive that year, with an average price of 1.84 dollars per pound.

Now I want to look at total 2021 prices by month. I'll filter to 2021 and then group by month.

```{r monthly_lastyear}
chs_mnths |>
  filter(yr == 2021) |> 
  group_by(month_nm) |> #group by both month and type
  summarize(
   average_monthly_price = mean(total_avg_price) 
) |> 
  arrange(average_monthly_price |> desc()
)
```

December 2021 saw the highest average total cheese price for that year, with an average price of 2.54 per pound.

### 2022 prices

Now I want to find the monthly average prices for this year, not adjusted for inflation.

```{r thisyear_price}
thisyear_price <- chs_mnths |> 
  filter(yr == 2022) |> 
  group_by(date) |> 
  summarize(avg_price = mean(total_avg_price))

thisyear_price
```

### Swiss prices by year

I want to look at just Swiss prices, so I'll filter by data to Swiss and look at means again.

```{r swiss_by_year}
chs_clean |> 
  filter(type == "SWISS") |> 
  group_by(yr) |> 
  summarize(
    yrly_adj_high_price = mean(adj_high_price),
    yrly_adj_low_price = mean(adj_low_price)
) |> 
  arrange(yrly_adj_high_price |> desc())
```

2014 had the highest average maximum Swiss cheese price after adjusting for inflation, with 3.84 dollars per pound in 2021 dollars.

The average high price for Swiss was around 90 cents lower in 2018, with an adjusted price of 2.95 dollars per pound.

### Muenster prices by year

I want to look at Muenster high and low prices, because it's my mom's favorite kind of cheese.

```{r cheddar}
chs_clean |> 
  filter(type == "MUENSTER") |> 
  group_by(yr) |> 
  summarize(
    yrly_adj_high_price = mean(adj_high_price),
    yrly_adj_low_price = mean(adj_low_price)
) |> 
  arrange(yr |> desc())
```

2014 saw the highest Muenster high prices after adjusting for inflation. That's the only year where the high price surpassed 3 dollars per pound. 

## Storage holdings analysis

### Which year had the highest cheese storage holdings, on average? Which had the lowest?

```{r storage}
storage_yrly <- storage_clean |> 
  filter(data_item == "CHEESE, NATURAL, COLD STORAGE, CHILLED - STOCKS, MEASURED IN LB") |> 
  group_by(year) |> 
  summarize(avg_yrly_holding = mean(value)) |> 
  arrange(year |> desc())

storage_yrly
```

2022 had the highest average cheese stocks, with over 1.48 billion pounds of cheese stored by the U.S. government. 

2001 had the lowest average storage holdings, with over 704 million.

### Which month, on average, sees the highest cheese storage holdings?

```{r storage_monthly}
storage_monthly <- storage_clean |> 
  filter(
    data_item == "CHEESE, NATURAL, COLD STORAGE, CHILLED - STOCKS, MEASURED IN LB"
) |> 
  group_by(month_use) |> 
  summarize(avg_storage = mean(value)) |> 
  arrange(avg_storage |> desc())

storage_monthly
```

The USDA stored the most cheese in July, storing an average with of 1.01 billion pounds of cheese during the summer month.

### Percent change in recent storage years

```{r percent_change}
storage_change <- storage_yrly |> 
  filter(year > 2013) |> 
  mutate(pct_change = (avg_yrly_holding/lead(avg_yrly_holding) - 1) * 100)

storage_change
```

Plotting percent change in recent years.

```{r plotchange}
percent_change_plot <- ggplot(storage_change, aes(x = year, y = pct_change)) +
  geom_col() +
  labs(title = "Percent change in U.S. Cheese Storage Holdings", subtitle = str_wrap("Cheese storage holdings have increased each year since 2014."), caption = "Chart by: Athena Hawkins, Source: USDA holdings data",
       x = "Year", y = "Percent change")

percent_change_plot
```


## My Visualizations

### 2022 Prices Plot (FOR MY STORY)

```{r thisyear_plot}
thisyear_plot <- ggplot(thisyear_price, aes(x = date, y = avg_price)) +
  geom_line() +
  geom_point() +
  labs(title = "Weekly U.S. cheese prices fluctuate throughout 2022", subtitle = str_wrap("Average prices per week for all domestic type cheeses peaked in June at over 50 cents higher than February's prices. According to an analysis of USDA market data, the last two weeks have seen falling prices."), caption = "Chart by: Athena Hawkins, Source: USDA Market Data",
       x = "Month", y = "Price per lb. (in 2022 dollars)")

thisyear_plot
```

### Storage over time (FOR MY STORY)

Now I want to plot storage over time by year. I'll use the tibble I made earlier, storage_yrly, for this. I want to use a bar chart so my audience can see that the storage holdings have nearly doubled.

```{r yrly_stor_cols}
yearly_storage_plot <- ggplot(storage_yrly, aes(x= year, y = avg_yrly_holding)) + 
  geom_col() +
  scale_y_continuous(labels = label_number(suffix = "B", scale = 1e-9)) + # billions
  theme_bw() +
  labs(title = "National Cheese Storage Holdings Surge Over Time",
       subtitle = str_wrap("The U.S. government has almost doubled its stockpile of surplus cheese since 2000. According to an analysis of USDA data, cheese storage reached nearly 1.5 billion pounds this year."),
       caption = "Chart by: Athena Hawkins, Source: USDA",
       x = "Year", y = "Average Cheese Stored (lbs)")

yearly_storage_plot
```

### Creating a facet plot

Now I want to create a facet plot for all cheese types over all the years in my data set, to give my readers a comprehensive look at prices.

```{r adj_facet_wrp}
facet_adj <- chs_with_avg |> 
  group_by(yr, type) |> 
  summarize(total_price_adj) |> 
  group_by(yr, type) |> 
  summarize(adj_yearly_price = mean(total_price_adj)) |>
  arrange()

facet_adj
```

Now I'll create the separate line plot for my inflation adjusted price data.

```{r types}
separate_plot <- ggplot(facet_adj, aes(x = yr, y = adj_yearly_price)) + 
  theme_bw() +
  geom_point(aes(color = type)) +
  geom_line(aes(color = type)) +
  labs(title = "Domestic type cheese prices over time (adjusted for inflation)",
       x = "Year", y = "Price per lb. (in 2021 dollars)")

separate_plot
```

### Inflation adjusted facet plot (FOR MY STORY)

Now I'll create the facet plot, which will separate the graphs by cheese type to show prices over time for each type of cheese. 

```{r facet_plot}
facet_plot <- separate_plot +
  facet_wrap(~ type) +
  theme_bw() +
  theme(legend.position = "none") +
  labs(title = "Domestic type cheese prices over time (adjusted for inflation)", subtitle = str_wrap("Cheese prices increased for all types in 2022, but over the last 20 years prices saw several ups and downs, according to an analysis of USDA market data. Prices hit a peak in 2014 due to higher demand for dairy exports from countries like China, according to the US Dairy Export Council."), caption = "Chart by: Athena Hawkins, Source: USDA market data",
       x = "Year", y = "Price per lb. (in 2021 dollars)")

facet_plot
```

### Facet Wrap All Cheeses, Not Adjusted

Just for fun, I want to see what the data would look like if I used my unadjusted numbers for the graphs.

```{r facet_unadj}
unadj <- chs_with_avg |> 
  group_by(yr, type) |> 
  summarize(total_avg_price) |> 
  group_by(yr, type) |> 
  summarize(unadj_price = mean(total_avg_price)) |>
  arrange()

unadj
```

I'll put together the separate line plot here. 

```{r}
unadj_plot <- ggplot(unadj, aes(x = yr, y = unadj_price)) + 
  geom_point(aes(color = type)) +
  geom_line(aes(color = type)) +
  labs(title = "U.S. Cheese Prices by Type (not adjusted for inflation)", 
       x = "Year", y = "Price of Domestic Cheese")

unadj_plot
```

Now I'll split it into individual plots.

```{r}
unadj_facet <- unadj_plot +
  facet_wrap(~ type) +
  theme(legend.position = "none")

unadj_facet
```

## Save plots

This will save the plots to my project folder as pdfs so I can look at the full scale.

```{r priceplot_save}
ggsave("thisyearprices.pdf", plot=thisyear_plot, width=12, height=8)
```

```{r facet_save}
ggsave("facetprices.pdf", plot=facet_plot, width=12, height=8)
```

```{r storage_save}
ggsave("yearlystorage.pdf", plot=yearly_storage_plot, width=12, height=8)
```

## Lede and nut graph draft

The U.S. government's stockpile of cheese reached nearly 1.5 billion pounds this year. The amount of cheese in cold storage across the country has more than doubled since 2001, according to an analysis USDA holdings data.

Government cheese holdings increased by over 2.8 percent this year, marking the eighth consecutive year in which the stockpile grew. This year's hike came amid rising prices for all types of domestic cheese, even after accounting for steep levels of inflation. But the federal government is no stranger to propping up dairy farmers amid economic uncertainty — their ever-growing cheese storage originated from another inflationary period in the 1970s.
